FEM Transition Models

### 2023-03-19 FEM_R/Loneliness_Social_Isolation/PAPER_WORKING_SCRIPT.Rmd
Duplicated the original repl population 10 times to have a bigger start point for cohort sims. Also added lots to the FEM_R scripts, including LY outcomes from detailed outputs, confidence intervals on plots, and visualisation of additional "realistic" interventions. Realistic interventions also added to the code, but still WIP to fine tune and get an interesting result. Added transition models for sight, hearing, and cognitive scores. Also transitioning these variables now

weighted.survey.means(s)

@@

print(mean.df)
```

@@

prepped.long %>% count(lnly)
```

@@

dead.alcstat <- dead %>%
  group_by(wave, lnly) %>%
  count() %>%
  rename(Deaths = 'n')
# 
# # replace NA with 'unknown'
# dead.alcstat$alcstat6 <- as.character(dead.alcstat$alcstat6)
# dead.alcstat[is.na(dead.alcstat)] <- 'Unknown'
# dead.alcstat$alcstat6 <- as.factor(dead.alcstat$alcstat6)
# 
# dead.alcstat
# 
# # Now plot deaths by alcstat6 at each wave
# ggplot(data = dead.alcstat, mapping = aes(x = wave, y = Deaths, fill = alcstat6)) +
#   geom_bar(stat = 'identity', position = 'dodge') +
#   labs(title = 'Deaths stratified by alcstat6 categories',
#        x = 'Wave',
#        y = 'Number of Deaths')
# 
# ## GOOD (not really big problems but the code works fine)
# # Now calculate the proportion of each group that die each wave
# 
# # First calculate n by group at each wave (including the unknowns)
# counts.subgroup <- test %>% 
#   group_by(wave, alcstat6) %>%
#   count() %>%
#   rename(Total_sample = 'n')
# 
# # Replace NA with 'unknown'
# counts.subgroup$alcstat6 <- as.character(counts.subgroup$alcstat6)
# counts.subgroup[is.na(counts.subgroup)] <- 'Unknown'
# counts.subgroup$alcstat6 <- as.factor(counts.subgroup$alcstat6)
# 
# # now merge total and deaths together
# counts <- merge(counts.subgroup, dead.alcstat, by = c('wave', 'alcstat6'), all = TRUE)
# 
# # finally, calculate proportion of deaths per subgroup
# counts$death_prop <- (counts$Deaths / counts$Total_sample) * 100
# 
# # Now plot death_prop by alcstat6 at each wave
# ggplot(data = counts, mapping = aes(x = wave, y = death_prop, fill = alcstat6)) +
#   geom_bar(stat = 'identity', position = 'dodge') +
#   labs(title = 'Proportion of deaths per wave within alcstat6 categories',
#        x = 'Wave',
#        y = 'Percentage of Sample')
# # one more without unknown
# ggplot(data = filter(counts, alcstat6 != 'Unknown'), mapping = aes(x = wave, y = death_prop, fill = alcstat6)) +
#   geom_bar(stat = 'identity', position = 'dodge') +
#   labs(title = 'Proportion of deaths per wave within alcstat6 categories',
#        x = 'Wave',
#        y = 'Percentage of Sample')
```

@@

# Cohort disease life years
diseaseVarLifeyear(cohort_full, 'cancre')
diseaseVarLifeyear(cohort_full, 'diabe')
diseaseVarLifeyear(cohort_full, 'hearte')
diseaseVarLifeyear(cohort_full, 'lunge')
diseaseVarLifeyear(cohort_full, 'hibpe')
diseaseVarLifeyear(cohort_full, 'stroke')
diseaseVarLifeyear(cohort_full, 'alzhe')
diseaseVarLifeyear(cohort_full, 'demene')
```

### Disability-free Life Years

### Disease-free Life Years

@@

#### Life Years
```{r}
coh.ly <- cohortLifeyear(cohort_full)
nolnly.ly <- cohortLifeyear(no.lonely)

@@

lifeyears <- rbind(ly.wide, disab.wide, disease.wide)
lifeyears['change'] <- lifeyears['intervention'] - lifeyears['cohort']
```

@@

ggsave(filename = 'ly_dfly_facet_L_S.png',
      plot = combined,
      path = path,
      width = 9,
      height = 6)
```

@@

```{r}
coh <- cohort

#base2 <- base %>% group_by(hhidpn, mcrep) %>%
#                  mutate(smoken_init = ())

# generate lifeyear for people who have not died
coh$lifeyear <- 1 - coh$died
# generate the number of repetitions for each person
coh$nreps <- max(coh$mcrep) + 1
# now generate the lifeyear intermediate (lifeyears per person)
coh2 <- coh %>% 
  group_by(hhidpn, year) %>%
  mutate(nreps = n())# %>%
  #ungroup() %>%
  #group_by(hhidpn) %>%
  #summarise(n_LY = sum(lifeyear),
  #       mean_LY = n_LY / nreps,
  #       n_DFLY = sum(nodisease),
  #       mean_DFLY = n_DFLY / nreps,
  #       two_yr_LY = mean_LY * 2,
  #       two_yr_DFLY = mean_DFLY * 2,
  #       weight_intermed = sum(weight),
  #       weight2 = weight_intermed / nreps)

#mean <- coh %>% summarise(LY = mean(mean_LY),
#                          DFLY = mean(mean_DFLY))
```

#### Gender
```{r}

@@

>>>>

weighted.survey.means(s)
```

@@

print(mean.df)

rm(design, means, mean.df, counts, s)
```

@@

prepped.long %>% count(lnly)

rm(test.out, dead.subgroup)
```

@@

dead.alcstat <- dead %>%
  group_by(wave, lnly) %>%
  count() %>%
  rename(Deaths = 'n')
# 
# # replace NA with 'unknown'
# dead.alcstat$alcstat6 <- as.character(dead.alcstat$alcstat6)
# dead.alcstat[is.na(dead.alcstat)] <- 'Unknown'
# dead.alcstat$alcstat6 <- as.factor(dead.alcstat$alcstat6)
# 
# dead.alcstat
# 
# # Now plot deaths by alcstat6 at each wave
# ggplot(data = dead.alcstat, mapping = aes(x = wave, y = Deaths, fill = alcstat6)) +
#   geom_bar(stat = 'identity', position = 'dodge') +
#   labs(title = 'Deaths stratified by alcstat6 categories',
#        x = 'Wave',
#        y = 'Number of Deaths')
# 
# ## GOOD (not really big problems but the code works fine)
# # Now calculate the proportion of each group that die each wave
# 
# # First calculate n by group at each wave (including the unknowns)
# counts.subgroup <- test %>% 
#   group_by(wave, alcstat6) %>%
#   count() %>%
#   rename(Total_sample = 'n')
# 
# # Replace NA with 'unknown'
# counts.subgroup$alcstat6 <- as.character(counts.subgroup$alcstat6)
# counts.subgroup[is.na(counts.subgroup)] <- 'Unknown'
# counts.subgroup$alcstat6 <- as.factor(counts.subgroup$alcstat6)
# 
# # now merge total and deaths together
# counts <- merge(counts.subgroup, dead.alcstat, by = c('wave', 'alcstat6'), all = TRUE)
# 
# # finally, calculate proportion of deaths per subgroup
# counts$death_prop <- (counts$Deaths / counts$Total_sample) * 100
# 
# # Now plot death_prop by alcstat6 at each wave
# ggplot(data = counts, mapping = aes(x = wave, y = death_prop, fill = alcstat6)) +
#   geom_bar(stat = 'identity', position = 'dodge') +
#   labs(title = 'Proportion of deaths per wave within alcstat6 categories',
#        x = 'Wave',
#        y = 'Percentage of Sample')
# # one more without unknown
# ggplot(data = filter(counts, alcstat6 != 'Unknown'), mapping = aes(x = wave, y = death_prop, fill = alcstat6)) +
#   geom_bar(stat = 'identity', position = 'dodge') +
#   labs(title = 'Proportion of deaths per wave within alcstat6 categories',
#        x = 'Wave',
#        y = 'Percentage of Sample')

rm(dead, dead.alcstat)
```

@@

# Cohort disease life years
diseaseVarLifeyear(cohort_full, 'cancre')
diseaseVarLifeyear(cohort_full, 'diabe')
diseaseVarLifeyear(cohort_full, 'hearte')
diseaseVarLifeyear(cohort_full, 'lunge')
diseaseVarLifeyear(cohort_full, 'hibpe')
diseaseVarLifeyear(cohort_full, 'stroke')
#diseaseVarLifeyear(cohort_full, 'alzhe')
diseaseVarLifeyear(cohort_full, 'demene')
```

@@

#### Life Years
```{r}
#coh.ly <- cohortLifeyear(cohort_full)
#nolnly.ly <- cohortLifeyear(no.lonely)

@@

lifeyears <- rbind(ly.wide, disab.wide, disease.wide)
lifeyears['change'] <- lifeyears['intervention'] - lifeyears['cohort']

rm(ly, ly.all, ly.wide, ly.lnly, ly.sociso, ly.lnly.sociso)
rm(disab.lnly, disab.sociso, disab.lnly.sociso, disab, disab.all, disab.wide)
rm(disease.lnly, disease.sociso, disease.lnly.sociso, disease, disease.all, disease.wide)
```

@@

ggsave(filename = 'ly_dfly_facet_L_S.png',
      plot = combined,
      path = path,
      width = 9,
      height = 6)

rm(combined, path)
```

### Using Detailed Output

```{r}
## Think it through - Steps:
# 1. Add a variable for lifeyears
  # Var needs to == 1 if not died and 0 if died
  # With this var we can calculate the total number of lifeyears for each person
# 2. Calculate lifeyears per person per repetition
  # Sum lifeyears per person
  # Divide sum lifeyears by number of reps
coh.test <- ly_outcomes_from_detailed_output(cohort)

ggplot(data = coh.test, aes(x = outcome, y = mean, fill = outcome)) +
  geom_bar(position = 'dodge', stat = 'identity') +
  geom_errorbar(aes(ymin = min, ymax = max),
                width = 0.2)

rm(coh.test)
```

#### Comparisons

First combine detailed outputs for each of the interventions we want to investigate.

```{r}
out.dir <- 'output/LNLY_SOCISO_TEST/'

cohort <- detailed_output_append(base.out.dir = out.dir,
                              scenario = 'ELSA_core_cohort')
nolnly <- detailed_output_append(base.out.dir = out.dir,
                              scenario = 'ELSA_nolnly_cohort')
nosociso <- detailed_output_append(base.out.dir = out.dir,
                              scenario = 'ELSA_nosociso_cohort')
both <- detailed_output_append(base.out.dir = out.dir,
                              scenario = 'ELSA_nolnly_nosociso_cohort')
```

```{r}
# Calculate LY outcomes for all scenarios
ly.coh <- ly_outcomes_from_detailed_output(cohort)
ly.lnly <- ly_outcomes_from_detailed_output(nolnly)
ly.sociso <- ly_outcomes_from_detailed_output(nosociso)
ly.lnly.sociso <- ly_outcomes_from_detailed_output(both)
# Add scenario column for combining
ly.coh$scenario <- 'Baseline'
ly.lnly$scenario <- 'No_Loneliness'
ly.sociso$scenario <- 'No_Social_Isolation'
ly.lnly.sociso$scenario <- 'Neither'

# Group into list and rbind all together
ly.list <- list(ly.coh, ly.lnly, ly.sociso, ly.lnly.sociso)
ly.all <- do.call(rbind, ly.list)

# arrange columns properly
ly.all <- ly.all %>% select(scenario, everything())

#rm(ly.coh, ly.lnly, ly.sociso, ly.lnly.sociso, ly.list)
```

Now we have a dataframe with the LY outcomes (and confidence intervals) for all scenarios. We need to reformat the dataset and plot these values with the confidence intervals clearly shown.

First do LifeYears.

```{r}
ly.all$scenario <- factor(ly.all$scenario, levels = c('Baseline', 'No_Loneliness', 'No_Social_Isolation', 'Neither'))
ly.all$outcome <- factor(ly.all$outcome, levels = c('LY', 'DiseaseFLY', 'DisabilityFLY'))

ly.ly <- ly.all %>% 
  filter(outcome == 'LY') %>%
  select(-outcome)

ggplot(data = ly.ly, aes(x = scenario, y = mean, fill = scenario)) +
  geom_bar(position = 'dodge', stat = 'identity') +
  geom_errorbar(aes(ymin = min, ymax = max),
                width = 0.2) +
  labs(title = 'Life Year Comparison between Intervention Scenarios') +
  xlab("Scenario") +
  ylab("Mean Life Years")

ly.disease <- ly.all %>%
  filter(outcome == 'DiseaseFLY') %>%
  select(-outcome)

ggplot(data = ly.disease, aes(x = scenario, y = mean, fill = scenario)) +
  geom_bar(position = 'dodge', stat = 'identity') +
  geom_errorbar(aes(ymin = min, ymax = max),
                width = 0.2) +
  labs(title = 'Disease Free Life Year Comparison between Intervention Scenarios') +
  xlab("Scenario") +
  ylab("Mean Disease Free Life Years")

ly.disability <- ly.all %>%
  filter(outcome == 'DisabilityFLY') %>%
  select(-outcome)

ggplot(data = ly.disability, aes(x = scenario, y = mean, fill = scenario)) +
  geom_bar(position = 'dodge', stat = 'identity') +
  geom_errorbar(aes(ymin = min, ymax = max),
                width = 0.2) +
  labs(title = 'Disability Free Life Year Comparison between Intervention Scenarios') +
  xlab("Scenario") +
  ylab("Mean Disability Free Life Years")
```

#### Change

```{r}
# Now do change plots starting with the ly.all object
ly.change <- ly.all %>%
  pivot_wider(names_from = 'scenario',
              values_from = c('mean', 'min', 'max', 'margin')) %>%
  mutate(L_change = mean_No_Loneliness - mean_Baseline,
         S_change = mean_No_Social_Isolation - mean_Baseline,
         LS_change = mean_Neither - mean_Baseline,
         L_min = L_change - margin_No_Loneliness,
         L_max = L_change + margin_No_Loneliness,
         S_min = S_change - margin_No_Social_Isolation,
         S_max = S_change + margin_No_Social_Isolation,
         LS_min = LS_change - margin_Neither,
         LS_max = LS_change + margin_Neither) %>%
  pivot_longer(cols = L_change:LS_max,
               names_to = 'scenario',
               values_to = 'LifeYears')

 # %>%
 #  mutate(L_min = L - margin_No_Loneliness,
 #         L_max = L + margin_No_Loneliness,
 #         S_min = S - margin_No_Social_Isolation,
 #         S_max = S + margin_No_Social_Isolation,
 #         LS_min = LS - margin_Neither,
 #         LS_max = LS + margin_Neither)

#ggplot(data = ly.change, aes(x = outcome, y = LifeYears, fill = scenario)) +
  #geom_col(position = 'dodge', stat = 'identity') +
  #geom_errorbar(ymin = (LifeYears - margin), ymax = (LifeYears + margin))
```

@@

#### Gender

@@

```{r}
test <- detailed_output_append(base.out.dir = 'output/COMPLETE/',
                              scenario = 'ELSA_core_cohort')
```

### 2023-03-19 FEM_R/Loneliness_Social_Isolation/transition_models_testing.Rmd
Duplicated the original repl population 10 times to have a bigger start point for cohort sims. Also added lots to the FEM_R scripts, including LY outcomes from detailed outputs, confidence intervals on plots, and visualisation of additional "realistic" interventions. Realistic interventions also added to the code, but still WIP to fine tune and get an interesting result. Added transition models for sight, hearing, and cognitive scores. Also transitioning these variables now

>>>>

# Sight

NOTE: Higher values of sight (self-rated eyesight) are worse. 1 == Excellent, 5 == Poor, 6 == Blind.

```{r}
trans.notDead$sight <- as.factor(trans.notDead$sight)

## MINIMAL
form.sight.min <- 'sight ~ male + scale(l2age65l) + scale(l2age6574) + scale(l2age75p)'
probit.sight.min <- polr(formula = form.sight.min,
                    data = trans.notDead,
                    Hess = TRUE,
                    na.action = na.omit)
print('Minimal model:')
summary(probit.sight.min)

## TEST
form.sight <- 'sight ~ male + white + hsless + college + scale(l2age65l) + scale(l2age6574) + scale(l2age75p) + l2catracte + l2logbmi_l30 + l2logbmi_30p + l2hchole + l2hibpe + l2alcfreq + l2smoken + l2smokev + l2stroke'
probit.sight <- polr(formula = form.sight,
                    data = trans.notDead,
                    Hess = TRUE,
                    na.action = na.omit)
print('Full model:')
summary(probit.sight)
```

AIC: 143800.44 : Initial
AIC: 128596.87 : Added alcfreq
AIC: 128133.34 : Added smoken smokev
AIC: 128039.81 : Added stroke

# Hearing

Using this paper - [Kiely et al.](https://academic.oup.com/biomedgerontology/article/67/9/997/513794?login=false).

```{r}
trans.notDead$hearing <- as.factor(trans.notDead$hearing)

## MINIMAL
form.hearing.min <- 'hearing ~ male + scale(l2age65l) + scale(l2age6574) + scale(l2age75p)'
probit.hearing.min <- polr(formula = form.hearing.min,
                    data = trans.notDead,
                    Hess = TRUE,
                    na.action = na.omit)
print('Minimal model:')
summary(probit.hearing.min)

## TEST
form.hearing <- 'hearing ~ male + white + hsless + college + scale(l2age65l) + scale(l2age6574) + scale(l2age75p) + l2logbmi_l30 + l2logbmi_30p + l2smoken + l2smokev + l2hibpe + l2diabe + l2stroke + l2sight'
probit.hearing <- polr(formula = form.hearing,
                    data = trans.notDead,
                    Hess = TRUE,
                    na.action = na.omit)
print('Full model:')
summary(probit.hearing)
```

AIC: 152711.79 : initial
